Basic concepts and tools for the Toki Pona minimalist language: Wordnet synsets; analysis, synthesis and syntax highlighting of texts

### 2017-11-19 article/article.tex
andado com artigo

\documentclass{article}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
	%pagebackref=true,
	pdfcreator={LaTeX with abnTeX2},
	pdfkeywords={abnt}{latex}{abntex}{USPSC}{trabalho acadêmico}, 
	colorlinks=true,       		% false: boxed links; true: colored links
	linkcolor=blue,          	% color of internal links
	citecolor=blue,        		% color of links to bibliography
	filecolor=magenta,      		% color of file links
	urlcolor=blue,
	allbordercolors=black,
	bookmarksdepth=4
}
\usepackage[utf8]{inputenc}
\usepackage{tipa}
% \usepackage{fontspec}
% \usepackage[mathletters]{ucs}
% \newcommand{\ttt}[1] {
% 	\texttt{<#1>}}
% \newcommand{\tttt}[1]{\texttt{#1}}

\begin{document}
\title{The Toki Pona Language:
an overview and some hacks}
% Anthropological computation
\author{Renato Fabbri\\
\texttt{renato.fabbri@gmail.com}\\
University of São Paulo,\\
Institute of Mathematical and Computer Sciences\\
São Carlos, SP, Brazil
}
\maketitle
\begin{abstract}

\end{abstract}

\section{Introduction}\label{intro}
Toki Pona is a minimalist conlang (constructed language)
with only 126 words.
% definition, historic, status (groups, speakers, etc).

\subsection{Resources on the web}
http://tokipona.net
\subsection{Historical note}

\subsection{Natural and constructed languages}
% conlangs, natural langs, programming langs, Lojban as an
% intermediary
\section{Overview of the language}\label{basics}
\subsection{Phonology}
Words in Toki Pona are written using only 14 letters:
\begin{itemize}
  \item Vowels a (open), e (mid front), o (mid back), i (close front),
    u (close back).
  \item Consonants j, k, l, m, n, p, s, t, w:
    \begin{itemize}
      \item Nasal: m (labial), n (coronal).
      \item Plosive: k (dorsal), p (labial), t (coronal).
      \item Fricative: s (coronal).
      \item Approximant: j (dorsal), l (coronal), w (labial).
    \end{itemize}
\end{itemize}

@@

\subsection{Syntax}
As in other natural languages, 
colloquial Toki Pona might have
incomplete sentences and deviate from the
norm.
The basic structure of sentences are in the form:
<subject> li <predicate> e <object>.
The li might be repeated to associate more than
one predicate to the subject.
The particle li is omitted if the subject is a simple mi (I or us)
or sina (you). A discussion about problems with this rule
and how I deal with them is in Appendix~\ref{mytoki}.

@@

Noun and verb phrases are built with the non-particle words.
The first word is the noun and phrase and subsequent words
qualify the noun or phrase.
The pi particle might be used to separate sequences of words
to be evaluated before the relations yield by pi:
As pi is often ill understood and used,
the following structures might be handy for newbies and as a
reference:
\begin{itemize}
  \item No pi, `word word word':
word $\leftarrow$ (qualifies 1) word $\leftarrow$ (qualifies 2)  word.
  \item One pi, 'word pi word word': word $\leftarrow$ (qualifies 2) [
      word $\leftarrow$ (qualifies 1)  word ].
  \item Two pi-s: `word pi word word word pi word word':
    word $\leftarrow$5 [word 2 word] 3 word $\leftarrow$4 word 1  word;
or:
    word $\leftarrow$5 [word 1 word] 2 word $\leftarrow$4 word 3  word.
\end{itemize}

Notes on the usage of pi:
\begin{itemize}
  \item In a sequence of words, without pi, the second word qualifies
    the first, the third word qualifies the phrase yield by the first
    two words, the fourth word qualifies the noun yield by the first
    three words and so on.
  \item It is redundant to use pi before the last word in a noun or
    verb phrase if there is no other pi, reason why it is most often
    omitted.
    Its use in this case is regarded as wrong~\cite{lipu,pije},
    but, as one might notice, it does not introduce any ambiguity.
  \item The book by jan Pije~\cite{kama} describes another use for pi:
    after li to mean possession, e.g. `soweli li pi sina' (your pet).
    This employment of pi might be regarded as correct, but are promptly written
    as a noun phrase (e.g. `soweli sina') and is not mentioned
    by the official book~\cite{tpLabg}.
\end{itemize}

All the words except the structural particles (li, e, la, pi)
are usable in noun and verb phrases.
Notice that the phrase expresses a noun in a noun phrase (subject or
object) or a verb (in the predicate).

At this point, the only missing syntax rule is related to
the prepositions: kepeken, lon, sama, tan, tawa.
They might appear at the end of noun phrases,
should be followed by  another noun phrase,
and require no particle. E.g.
`toki tan jan Pije li pana e sona tawa mi'.

Other particles are:
\begin{itemize}
  \item a or kin, emphasis.
  \item o, vocative or imperative ('jan lukin sitelen o, li wawa')
  \item taso, means however or 'only' if adjective.
  \item anu, en: 'or' and 'and'. Used for nouns in nous phrases.
    For repetition of verbs, repeat li.
    For object nouns, repeat e.
    If the noun is complementing a preposition (tawa, lon),
    one might repeat the preposition.
    As Toki Pona is a recent language, and is able to cope with
    variation due to its simplicity, I would advocate for
    using en and anu wherever there is no ambiguity.
  \item nanpa, denotes numbering.
  \item seme, for questions, used next to the thing being asked for.
    'Why?' might be expressed as
    'seme la sina pana e moku lon sewi'.
  \item mu, for animal noises. For me it is not a particle, as in the
    official dictionary, but a noun.
    I also like to use it as a verb:
    'mi pakala e luka. mu mute.
\end{itemize}

@@

As far as I can see, one should regard
the particles li, e, pi, and la
and punctuation.
The other tokens of the vocabulary
might be used in any of the remaining positions.
Detection:
\begin{itemize}
  \item Noun: the first word in a noun phrase.
    After an 'e' and after a pi,
    The first word in the sentence if
    sentence does not start by the verb.
    Might be in the position of the verb
    if the sentence has no object.
  \item Adjective: second word on after an e and after
    Second word on in the subject phrase if present.
  \item Verb: after a li, mi or sina.
    If there is no object, the verb position
    is often a noun.
  \item After a preposition,
    there can be a noun phrase, a verb phrase
    or nothing.
  \item Notice that there is ambiguity in the structure
    introduced by the omission of li after mi and sina.
    Also, when there is no object, a noun or a verb
    or an adjective might be in the verb position
    if there is no object.
    The prepositional complement is also not defined.
    (mi moku tawa pali, tawa tomo).
    So, these are sources of syntactic ambiguities
    in Toki Pona.
    They might be solved or minimized by using the semantics
    of the words.
    One preliminary effort in this direction might be
    using the classes in the official dictionary to resolve
    ambiguities whenever possible.
    This solution is not optimal in correct POS tagging,
    and does not solve all possible ambiguities
    (there are words classified as nouns and adjectives,
    adjectives and verbs, nouns and verbs, particle and verb).

@@

\subsection{Further notes}
The only synonyms on Toki Pona are:
a and kin; lukin and oko;
sin or kamako;
ale or ali.

@@

% my view
% considerations about the official book and jan Pije
% pictorial forms
\subsection{Main references for the language}
\begin{itemize}
  \item The official book is ``Toki Pona: The Language of Good''
    and is authored by Sonja Lang, the creator of the language.
  \item The book ``o kama sona e toki pona!'', from jan Pije,
    is the other main reference for the language~\cite{kama}.
  \item \cite{Wikipesija}.
\end{itemize}

\section{Analysis and hacks for the language}\label{hacks}

\section{Statistics of the vocabulary}
In~\cite{pije} there is statistics about Toki Pona corpus.
This section focuses on the statistics of the vocabulary
and syntactic rules:
the letters, phonemes, word sizes,
possible combinations for words and sentences.
The statistics are in Appendix~\cite{listings},
and the next paragraph is an overview.
Python scripts were used to obtain the measurements
and are available at~\cite{tokipona}.

@@

\section{Synthesis of text}
The same package~\cite{tokipona}
has capabilities for synthesizing text.
Noun, verb and prepositional phrases,
sentences.
It also aims at making larger scale texts
by keeping a record of the used words and structures (context)
and using stylistic outlines for poems and short narratives.

@@

\section{Syntax highlighting}
The same package~\cite{tokipona}
has a Vim syntax highlighting plugin
for Toki Pona.
Instructions for installing and using
the syntax highlighting is at~\cite{tokipona}.

@@

@@

\section{Conclusions and further work}\label{conc}
\begin{itemize}
  \item Relate Toki Pona to Wordnet: should one Toki Pona word
    be related to more then one synset of the English language?
  \item Understand how the corpus is gathered in \url{tokipona.net}.
  \item Know about previously existing words that were used for Toki Pona
    (e.g. suno and suwi might come from sun and sweet),
    and about the reasons that lead Sonja (and maybe other people)
    to choose the 14 letters and the syllable structure.
    This might require a dedicated communication with the
    speaker community and the documentation authors.
  \item Corpus-based analysis.
  \item Publication of original texts and translations.
  \item Make an article written in Toki Pona.
    I wrote Section~\ref{finalToki}, and believe that
    a summary in both English and Toki Pona for
    facilitating the acquisition of context,
    a reasonable article on some scientific topics is
    possible.
    I first conceived something around complexity, statistics, physics, or computer science.
    But one possibility is to write about linguistics, philosophy, literature
    or psychology with the partners I write in English.
    I can start a draft,
    they might learn the language in a few hours (with or without my help),
    to contribute, and we can write a short paper.
  \item Enhance the synthesis of text to yield better contextualized text
    and stylistic traces, such as for poems and short stories.
    Give the user the ability to choose the sentences (generate randomly according to
    previously written or given text,
    some rules input by the user, the package and the language guidelines,
    outputs to the screen and asks to keep and discard).
  \item Enhance the syntax highlighting implementation described in Section~\ref{high}.
    It uses only syntactic cues to choose POS tags,
    which might be overcome by using n-grams and
    further techniques from Natural Language Processing~\cite{POS}.
\end{itemize}

\subsection*{Acknowledgments}
FAPESP (project 2017/05838-3); Vim developers and documentation maintainers;
Vim user community.

\appendix
\section{My usage of Toki Pona}\label{mytoki}
I use the standard sounds, but often use [z] for s.
I often translate texts to Toki Pona (e.g. biblical excepts)
and create new texts as poems and short stories.
Most of them are in~\cite{tokisona}.
I omit the li particle after subjects sina and mi,
in accordance with the norm,
but sometimes I use them when there are many predicates.
E.g. sina li wawa li pimeja li lukin pona li moku e kasi mute.
In such cases, the first li is sometimes omitted.
Also, sometimes I use li before mi and sina where I find
that there is unwanted ambiguity, e.g.
sina moku pona e jan 
(might be sina li moku pona e jan or sina moku li pona e jan).

@@

\input{pos}

>>>>

\documentclass{article}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
	%pagebackref=true,
	pdfcreator={LaTeX with abnTeX2},
	pdfkeywords={abnt}{latex}{abntex}{USPSC}{trabalho acadêmico}, 
	colorlinks=true,       		% false: boxed links; true: colored links
	linkcolor=blue,          	% color of internal links
	citecolor=blue,        		% color of links to bibliography
	filecolor=magenta,      		% color of file links
	urlcolor=blue,
	allbordercolors=black,
	bookmarksdepth=4
}
\usepackage[utf8]{inputenc}
\usepackage{tipa}
\newcommand{\ttt}[1] {
	\texttt{<#1>}}
\newcommand{\tttt}[1]{\texttt{#1}}
% \usepackage{fontspec}
% \usepackage[mathletters]{ucs}
% \newcommand{\ttt}[1] {
% 	\texttt{<#1>}}
% \newcommand{\tttt}[1]{\texttt{#1}}

\begin{document}
\title{Basic concepts and tools for the Toki Pona minimalist language:
Wordnet synsets; analysis, synthesis and syntax highlighting of texts}
% Anthropological computation
\author{Renato Fabbri\\
\texttt{renato.fabbri@gmail.com}\\
University of São Paulo,\\
Institute of Mathematical and Computer Sciences\\
São Carlos, SP, Brazil
}
\maketitle
\begin{abstract}
  A minimalist constructed language (conlang)
  is useful for experiments and is comfortable for making tools.
  The Toki Pona conlang is minimalist both in the vocabulary
  (only 14 letters, X phonemes, and 120 words + 4 synonyms)
  and in the $\approx10$ syntax rules.
  The language is useful for being a used and somewhat established
  minimalist conlang with at least hundreds of speakers.
  In this article, we describe current concepts and resources
  for Toki Pona,
  and make available Python scripted routines for
  analysis and synthesis of texts and the specification of syntax highlighting schemes.
  We focus on the analysis of the basic vocabulary,
  as corpus analyses were found in~\cite{corpus}.
  The synthesis is based on the basic sentence template,
  relates to context by keeping track of used words,
  and renders larger texts by using a fixed number of phonemes (e.g. for poems)
  and number of sentences, words and letters (e.g. for paragraphs).
  Syntax highlighting is performed in 24bit true colors, currently only
  with the Vim text editor.
  Word colors reflect morphosyntactic classes given in the official dictionary
  and different solutions to necessary and arbitrary choices are described and implemented.
  In summary, this text holds potentially novel conceptualizations about,
  and tools and results in analyzing, synthesizing or syntax highlighting the
  Toki Pona language.
  % subj (li), obj (e), prepositions, qualification (word concatenation), 
  % sent (la), particles, mi-sina, pi, punctuation is freestyle, pre-verb,
\end{abstract}
{\bf keywords:} Natural Language Processing, Textual synthesis, Syntax highlighting, Wordnet, Toki Pona

\section{Introduction}\label{intro}
Toki Pona is a minimalist conlang (constructed language)
with only 124 words (120 without the synonyms).
Therefore, the concepts are usually very general and different,
and, without context, the words are rarely related through
meronymy and hyponymy.
Such a linguistic setting is desired because of the simplicity
which entails easier e.g. learning and tool making.
Another reason why the minimalistic language design is compelling
is the study and harnessing of the strong and weak forms of
the Sapir-Whorf hypothesis (linguistic relativity),
i.e. that language influences or dictates ones thought
and world experience.
Accordingly, one uses a conlang as a thinking tool (or platform)
or to make experiments about the influence language has on
the thoughts of the `speaker' (also writer and reader).
Toki Pona is often described as a tool to meditate,
to simplify the thinking processes, and as a way to modify
the mood and impressions about the world.~\cite{tp1,tp2,tp3,tp4}
In~\cite{interview}, Sonja Lang (the creator of Toki Pona),
describes that she has seen the language been used successfuly
in the context of management, creation of texts, legal texts,
etc.

In this article, we present a conceptual overview of the language,
which is fit both to the newcomer and to the expert.
We also summarize
texts and software gadgets
for Toki Pona that
were developed by us and by other Toki Pona users.
Most importantly, the conceptual overview is considerably different
from what we found in the literature~\cite{pu,pije,memrise},
with emphasis on simplicity and flexibility,
and the software routines yield analyses, syntheses
and highlighting Toki Pona what was thought to be complementary
to the software currently available.

Next subsections hold a description of the general resources available,
a historical note, and some words about natural and constructed languages.
Section~\ref{basic} describes Toki Pona's phonology and syntax.
Section~\ref{hacks} presents the software routines we made available
and immediate results, such as listings and statistics of words,
poems and short stories, and coloring schemes.
Conclusions and further work are in Section~\ref{conc}.
Appendix~\ref{mytoki} holds considerations about my usage of Toki Pona,
with thoughts of rule breaking and potentially new conlangs.
Appendix~\ref{ftp} holds final words in Toki Pona.

% definition, historic, status (groups, speakers, etc).

\subsection{Resources on Toki Pona}
One might organize current resources for the Toki Pona
language in: references and learning material, corpus,
websites, interaction groups (here users talk a post texts and comments), and software gadgets.
The main references of the language are:
  the official book ``Toki Pona: The Language of Good'',
    authored by Sonja Lang, the creator of the language;
the online book ``o kama sona e toki pona!'',
from jan Pije~\cite{kama}.
For a more comprehensive view of the resources available
for the user, we suggest following the links 
from~\cite{wikiToki}.
Section~\cite{third} holds a summary of what we found in software
gadgets related to Toki Pona.

\subsection{Historical note}
Toki Pona was developed as an internal and personal language
by Sonja Lang~\cite{interview}.
It was released as a draft in 2001 and in 2007 some documents
reported it to have a few hundred speakers.
The English official book~\cite{lipu} was released only in 2014.
In 2016, a version of the official book was released in French.
Nowadays, one finds a number of texts about Toki Pona and written
in it, and uses of Toki Pona for artificial intelligence~\cite{roila}
and software tools~\cite{tptoll1,tptool2,tptool3},
in social platforms such as Facebook groups~\cite{fbtp,fbtpt},
microblogging~\cite{tokilili}, Telegram~\cite{telegram} and IRC~\cite{irc}.

\subsection{Natural and constructed and artificial languages}
A language one uses (or might use) to communicate by speaking
and writing is called `natural language'.
A `constructed language' is a natural language built by someone or
a group, such as Esperanto, Toki Pona, and Lojban.
An artificial language is a language yield by artificial agents,
such as in AI routines, and might be considered
within `cultural evolution' studies.
Formal languages are defined by tokens and rules to
operate them, they span from computer programming languages
to math and formal models for natural languages.
Lojban is a constructed language with the purpose of being
an artificial language because it is more axiomatic.

Constructed languages are in some traditions called artificial
languages, but creators most often prefer to use the term `standardized'
or 'planned' language with the argument that the conlang
is rooted on natural languages, and 'artificial' is misleading~\cite{conlanWikip}.
The preferred terms seem to be planned or constructed languages
or conlang.
The construction of languages is called glossopoeia.
Toki Pona is a conlang which might be classified
as engineered for experiments, meditation and philosophy;
and suitable for auxiliary international language and
as an artistic language. 
% taxonomy, vocabulary, ontology, terminology,

% conlangs, natural langs, programming langs, Lojban as an
% intermediary
% non-verbal languages, musical corporal language, 
% non-verbal languages confront Sapir-Whorf hypothesis.

\section{Overview of the language}\label{basics}
This section describes very succinctly the formation of
words and sentences in the Toki Pona language.
It should enable a newcomer to grasp the essentials
of the language and the experienced to
acquire new insights.
Furthermore, it is a solid reference of the syntax and phonological
rules.

\subsection{Phonology}
Words in Toki Pona are written using only 14 letters:
\begin{itemize}
  \item Vowels a (open), e (mid front), o (mid back), i (close front),
    u (close back).
  \item Consonants j, k, l, m, n, p, s, t, w:
    \begin{itemize}
      \item Nasal: m (labial), n (coronal).
      \item Plosive: k (dorsal), p (labial), t (coronal).
      \item Fricative: s (coronal).
      \item Approximant: j (dorsal), l (coronal), w (labial).
    \end{itemize}
\end{itemize}

@@

\subsection{Syntax}
\subsubsection{Fundamental notions}
As in other natural languages, 
colloquial Toki Pona might have
incomplete sentences and deviate from the
norm.
The basic structure of sentences is:
`subject' (Noun) li `predicate' (Verb) e `object' (Noun).
The li might be repeated to associate more than
one predicate to the subject.
The particle li is omitted if the subject is a simple mi (I or us)
or sina (you). A discussion about issues (problems)
yield by this rule
and how I deal with them is in Appendix~\ref{mytoki}.

@@

Noun and verb phrases are (usually) built with the non-particle words.
The first word is the noun or verb and subsequent words
qualify the noun or verb.
The pi particle might be used to separate sequences of words
to be evaluated before the relation yield by pi.
As pi is often ill understood and used,
the following structures might be handy for newbies and as a
reference:
\begin{itemize}
  \item No pi, `word word word':
word $\leftarrow$ (qualifies 1) word $\leftarrow$ (qualifies 2)  word.
  \item One pi, 'word pi word word': word $\leftarrow$ (qualifies 2) [
      word $\leftarrow$ (qualifies 1)  word ].
  \item Two pi-s: `word pi word word word pi word word':
    word $\leftarrow$5 [word 2 word] 3 word $\leftarrow$4 word 1  word;
or:
    word $\leftarrow$5 [word 1 word] 2 word $\leftarrow$4 word 3  word.
\end{itemize}

Notes on the usage of pi:
\begin{itemize}
  \item In a sequence of words, without pi, the second word qualifies
    the first, the third word qualifies the phrase yield by the first
    two words, the fourth word qualifies the noun yield by the first
    three words and so on.
  \item It is redundant to use pi before the last word in a noun or
    verb phrase if there is no other pi, reason why it is most often
    omitted.
    Its use in this case is regarded as wrong~\cite{lipu,pije},
    but, as one might notice, it does not add (much) information
    through syntax (order of qualifies is conserved).
    It adds as an emphasis because of greater length of the written segment, as a preparation: 'jan lili pi mama' (mommy's child).
    Also used in~\cite{akesiWawa}.
  \item The book by jan Pije~\cite{kama} describes another use for pi:
    after li to mean possession, e.g. `soweli li pi sina' (your pet).
    This employment of pi might be regarded as correct, but are promptly written
    as a noun phrase (e.g. `soweli sina') and is not mentioned
    by the official book~\cite{tpLabg}.
\end{itemize}

All the words except the structural particles (li, e, la, pi)
are usable in any position in noun and verb phrases,
(i.e. 116 words).
Notice that the phrase expresses a noun in a noun phrase (subject or
object) or a verb (in the predicate).
And that the first word of the phrase is the noun or verb,
and that subsequent words are adjectives or adverbs.
The pi and other particles restarts the phrase to
first word being a noun.

At this point, the only missing syntax rule is related to
the prepositions: kepeken, lon, sama, tan, tawa.
They might appear at the end of phrases,
should be followed by another phrase,
and require no particle. E.g.
`toki *tan* jan Pije li pana e sona *tawa* mi'.
Because prepositions can be used as adjectives
or nouns or verbs, adverbs, etc, and what follows
a preposition complement might also be of any morphosyntactic class,
they are core sources of ambiguity.
three:
* preposition
* absence of li on sina and mi
* Incomplete sentences
(ona li) tawa lon kon e ilo pi suli mute

\subsubsection{Particles}
Beyond the structural particles (li, e, la, pi) presented
in last section, other particles are:

\begin{itemize}
  \item a or kin, emphasis.
  \item o, vocative or imperative ('jan lukin sitelen o, li wawa')
  \item taso, means however as sentence or 'only' if adjective.
  \item anu, en: 'or' and 'and'. Used for nouns in nous phrases.
    For repetition of verbs, repeat li.
    For object nouns, repeat e.
    If the noun is complementing a preposition (tawa, lon),
    one might repeat the preposition.
    As Toki Pona is a recent language, and is able to cope with
    variation due to its simplicity, I would advocate for
    using en and anu wherever there is no ambiguity.
    E.g.: 'mi tawa anu moku lon tenpo lili?'
    In the official documentation that is not described and
    might be regarded as wrong in strict canonical.
  \item nanpa, denotes numbering.
  \item seme, for questions, used next to the thing being asked for.
    'Why?' might be expressed as
    'seme la sina pana e moku lon sewi'.
  \item mu, for animal noises. For me it is not a particle, as in the
    official dictionary, but a noun.
    I also like to use it as a verb:
    'mi pakala e luka. mu mute.'
\end{itemize}

@@

As far as I can see, one should regard
the particles li, e, pi, and la
and punctuation.
The other tokens of the vocabulary
might be used in any of the remaining positions.

\subsubsection{Convention for recognizing POS tags by speaker and machine}

\begin{itemize}
  \item Noun: the first word in a noun phrase.
    After an 'e' and after a pi,
    The first word in the sentence if
    sentence does not start by the verb.
    Might be in the position of the verb
    if the sentence has no object.
  \item Adjective: second word on after an e and after
    Second word on in the subject phrase if present.
  \item Verb: after a li, mi or sina.
    If there is no object, the verb position
    is often a noun.
  \item After a preposition,
    there can be a noun phrase, a verb phrase
    or nothing.
  \item Notice that there is ambiguity in the structure
    introduced by the omission of li after mi and sina.
    Also, when there is no object, a noun or a verb
    or an adjective might be in the verb position
    if there is no object.
    The prepositional complement is also not defined.
    (mi moku tawa pali, tawa tomo).
    So, these are sources of syntactic ambiguities
    in Toki Pona.
    They might be solved or minimized by using the semantics
    of the words.
    One preliminary effort in this direction might be
    using the classes in the official dictionary to resolve
    ambiguities whenever possible.
    This solution is not optimal in correct POS tagging,
    and does not solve all possible ambiguities
    (there are words classified as nouns and adjectives,
    adjectives and verbs, nouns and verbs, particle and verb).

@@

\subsubsection{Further notes}
The only synonyms on Toki Pona are:
a and kin; lukin and oko;
sin or kamako;
ale or ali.

@@

% my view
% considerations about the official book and jan Pije
% pictorial forms
\section{Software for analysis and synthesis and for syntax highlighting}\label{hacks}
In this section,
we describe software, natural language, audiovisual and numeric results
on four directions:
1)
the analysis of the Toki Pona language through statistics
obtained by processing the dictionary;
2)
the synthesis of sentences, poems and short stories
in Toki Pona;
3)
syntax highlighting for Toki Pona,
including fine-tuning and theoretical
considerations;
4)
initial considerations about the WordNet of
Toki Pona words.
Corpus analysis was already found in~\cite{}.

\subsection{Statistics of the vocabulary}
In~\cite{pije} there is statistics about Toki Pona corpus.
This section focuses on the statistics of the vocabulary
and syntactic rules:
the letters, phonemes, word sizes,
possible combinations for words and sentences.
The statistics are in Appendix~\cite{listings},
and the next paragraph is an overview.
Python scripts were used to obtain the measurements
and are available at~\cite{tokipona}.

@@

\subsection{Synthesis of text}
number of elisoes

The same package~\cite{tokipona}
has capabilities for synthesizing text.
Noun, verb and prepositional phrases,
sentences.
It also aims at making larger scale texts
by keeping a record of the used words and structures (context)
and using stylistic outlines for poems and short narratives.

@@

\subsection{Syntax highlighting}
The same package~\cite{tokipona}
has a Vim syntax highlighting plugin
for Toki Pona.
Instructions for installing and using
the syntax highlighting is at~\cite{tokipona}.

@@

The tokipona.vim syntax file has an association of toki pona
words with highlighting groups.
It also holds association of these groups to other groups
to use their color settings.
Therefore, one is able to use various color schemes
with syntax-highlighted Toki Pona texts.
A Vim user might run \tttt{:colorscheme blue},
\tttt{:colorscheme solarize}, \tttt{:colorscheme gruvbox}
\tttt{:colorscheme elflord} to see the same text colored
with different colorschemes.
To interfere directly on the colors chosen,
\tttt{:highlight Normal guifg=\#00000 guibg=\#0000ff}
will change the standard foreground (text) color to pure black
and background to pure blue.
See \tttt{:h gui-colors} and \tttt{:h highlight}
for the way you might edit colors directly.
Useful commands are given as commented lines
in the end of the syntax file.

\subsubsection{Advanced syntax highlighting considerations}
Standard guidelines for syntax highlighting
depend heavily on cultural and use factors
(https://softwareengineering.stackexchange.com/questions/89936/syntax-highlighting-color-scheme-studies),
although there are projects such as Solarized~\cite{solazired}
(http://ethanschoonover.com/solarized)
which try to give general solutions for some contexts.
Standards guidelines for syntax highlighting were not found,
and might be driven from data visualization considerations.
Most evident examples of such guidelines are:
\begin{itemize}
  \item The use of blue and other high frequency colors for
  background.
  This is extended to other high frequency (small wavelength)
  colors such as violet.
  \item The avoidance of colors of blue and other high frequency
  colors in small objects, such as letters and words.
  \item Explore simplicity and elegance through the use of discrete
  coloring schemes. Most of the tokens should be preferably of
  the same color.
  One might use a power-law to mimic natural occurring phenomena
  and perception. Or an exponential to map parameters to perceptual
  stimuli.
  \item You might always wish to maximize contrast,
  although taste and less consative combinations might
  also dictate choices.
  \item Stipulate axes of parameters with which to set colors.
  Wavelength or frequency is an obvious axis given the considerations
  above.
  Other parameters are hue, transparency etc.

  two types of colorschemes: those with dark background,
  which are more commode at first; and those with a light background,
  which are usually impressive (and even annoying) at first,
  but the eye gets used to it and it keeps you more stimulated.
  The blue color seems to be specially related to physiological
  stimulation of the body by color~\cite{blue,blue}.
  \item Consider that the user might stay many hours (and we often do)
  at a text editor, and that the colors and formats involved are
  prone to entail a considerable effect in the body and mental
  activity.
  \item Syntax sonification: consider mapping the textual structures
  to sound (e.g. counting of tokens or context as inside a function or
  class or over or around a variable of conditional or loop.
  E.g. through MASSA.
  \item Mode for tuning syntax highlighting with keyboard shortcuts
  are used to tune individual colors from background and foreground
  colors and groups.
  \item How would these techniques be used for coloring Toki Pona?
  Coloring of keywords that reflects the dictionary classification
  of words are not in a power law distribution.
  Maybe better described by a half-normal or by
  overlapping normal distributions over the three peaks ($\sim35-58,
  12-18, 1-5 occurrences$).
  \item Coloring for poems: division by syllables, words, POS tag,
    total number of syllables or letters of words.
\end{itemize}

\url{http://pygments.org/docs/}

\subsection{Wordnet Toki Pona}\label{wn}
exception: jan (people)
is a hyponym of soweli (mammal),
kili is hyponym of kasi,
(walo, pimeja, jelo, loje, laso) are hyponyms of kule.
But otherwise such relations hold:
e.g. luka (as hand) is e meronym of sijelo (body).

For a first achievement of a Toki Pona Wordnet,
each of the Toki Pona words in the dictionary
were related to Wordnet synsets
through the English lemmas.
All synsets related to the lemma with the POS tag
given in the dictionary are added.
Another immediate possibility would be to
relate each Toki Pona word to the synsets
related to all the words bounded by a semicolon.
Another possibility, which I find more consistent
but also more controversial,
is to choose all synsets related to the Wnglish
words given that the Toki Pona words can be used
in many roles.
WordNet only contains nouns, adjectives, verbs, and adverbs.
Thus, particles and prepositions were not considered.
Numbers were considered adjectives.
Words presented as adjectives in the dictionary
were considered adjectives and adverbs.

The total number of different synsets is 2493 in a very conservative
counting: discarding prepositions and particles,
only considering the synsets which have the POS tag
given in the dictionary.
If all the synsets (that have any POS tag)
related to a lemma are considered, 
then Toki Pona vocabulary reaches 3961 synsets.
The prepositions are particularly missing in these
countings because e.g. tawa or lon are used also
as ordinary words.
If synsets for them are considered those of the English
lemmas with any POS tag, the vocabulary reaches 4052
synsets.
The total number of synsets in Wordnet is X.
Total number of synsets counting neighbors.
Depth of TP synsets.
Using such features to devise other conlangs.

@@

\section{Conclusions and further work}\label{conc}
\begin{itemize}
  \item Relate Toki Pona to Wordnet: should one Toki Pona word
    be related to more then one synset of the English language?
    An immediate implementation would be linking each Toki Pona word
    to the English words in the official vocabulary.
    A Toki Pona synset would be related to all the synsets of each English
    words regarded as lemmas (sequence of characters or sound).
  \item Translators. This Toki Pona Wordnet would be useful e.g. for naive translators
    of words and text.
    Enhancements of such translators would use POS tagging and templates
    in English and Toki Pona and/or corpus and further statistical methods.
    An efficiency of Toki Pona can be studied in terms of the coverage of
    the Wordnet cloud.
  \item Conlanging. In fact, I plan to use Wordnet for making new conlangs
    (conlanging), and insights derived from Toki Pona might be used for that end.
  \item Understand how the corpus is gathered in \url{tokipona.net}.
  \item Know about previously existing words that were used for Toki Pona
    (e.g. suno and suwi might come from sun and sweet),
    and about the reasons that lead Sonja (and maybe other people)
    to choose the 14 letters and the syllable structure.
    This might require a dedicated communication with the
    speaker community and the documentation authors.
  \item Corpus-based analysis.
  \item Publication of original texts and translations.
  \item Make an article written in Toki Pona.
    I wrote Section~\ref{finalToki}, and believe that
    a summary in both English and Toki Pona for
    facilitating the acquisition of context,
    a reasonable article on some scientific topics is
    possible.
    I first conceived something around complexity, statistics, physics, or computer science.
    But one possibility is to write about linguistics, philosophy, literature
    or psychology with the partners I write in English.
    I can start a draft,
    qhey might learn the language in a few hours (with or without my help),
    to contribute, and we can write a short paper.
  \item Enhance the synthesis of text to yield better contextualized text
    and stylistic traces, such as for poems and short stories.
    Give the user the ability to choose the sentences (generate randomly according to
    previously written or given text,
    some rules input by the user, the package and the language guidelines,
    outputs to the screen and asks to keep and discard).
  \item Enhance the syntax highlighting implementation described in Section~\ref{high}.
    It uses only syntactic cues to choose POS tags,
    which might be overcome by using n-grams and
    further techniques from Natural Language Processing~\cite{POS}.
  \item We are particularly interested in software tools,
    and one possibility is to use a conlang to describe software
    routines.
    In this sense, it seems profitable
    to understand what `non-ambiguous' means in Lojban,
    and in what sense one is able to compile and parse Lojban.
    This study will probably yield considerations about
    parsing Toki Pona files.
  \item Conlang idealized for describing data, programs,
    scientific writings, creative writing
    (exploring the thinking process).
    Different modes that might be specified by a section header or tag.
\end{itemize}

\subsection*{Acknowledgments}
FAPESP (project 2017/05838-3);
Vim, Python, and Toki Pona authors, developers and literature maintainers;
user and speaker communities.

\appendix
\section{My usage of Toki Pona}\label{mytoki}

I use the standard sounds, but often use [z] for s.
I often translate texts to Toki Pona (e.g. biblical excepts)
and create new texts as poems and short stories.
Most of them are in~\cite{tokisona}.
I omit the li particle after subjects sina and mi,
in accordance with the norm,
but sometimes I use them when there are many predicates.
E.g. sina li wawa li pimeja li lukin pona li moku e kasi mute.
In such cases, the first li is sometimes omitted.
Also, sometimes I use li before mi and sina where I find
that there is unwanted ambiguity, e.g.
sina moku pona e jan 
(might be sina li moku pona e jan or sina moku li pona e jan).

@@

pali moku is addiction.
noka and ampa; luka, suno, sike and lawa;
I tend to group these things, as pali and pana.
These clusters of meaning hint me that an even simpler
language is possible and maybe more optimized to the
core goals of Toki Pona.
Also, ambiguities introduced by omission of li
and absence of a token to denote preposition,
suggest to me the possibility of a syntax that is uniquely
parsed.

I am tending to think on a language with the same syntax of toki pona:
<sentence> la <sentence = subject + predicate + object + preposition complement>
but always using the syntax separators,
repeating them when the function in performed in the smallest scale:
'(toki pi toki pona) pi jan sona' (language in toki pona and of intellectuals) vs
'toki pi (toki pona pi pi jan sona)' (language in toki pona of intelectuals).
The keyword for preposition complement is the catch.
An 'a' seems good, but it is taken in Toki Pona.
Maybe use it as the preposition particle and use ha or he
instead of Toki Pona a.
In such a setting,
one might enable pi li e a (concept-qualifier, subject-predicate,
predicate-object, concept-predicate)
in any of the terms of the template:
<sentence> la <sentence = subject + predicate + object + preposition complement>
Between sentences one would use:
la pi, la li, la e, la a, and la la as normally.
This is a very fractal conlang proposal.
Maybe have a way to distinct between noun, qualifiers (adjectives/adverbs) and verb terms,
and accept any of them for subject, predicate, object, predicate,
slots. Or assume a part of speech for template slot or for first word in slot.
